\documentclass[a4paper]{article}

\usepackage{fancyvrb}
\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small}
\DefineVerbatimEnvironment{example}{Verbatim}{fontsize=\small}
\newcommand{\ignore}[1]{}

\title{Monadic Combinators for Subjective Logic}
\author{Bryan St. Amour}

\begin{document}

\maketitle

\section{Introduction}

Subjective Logic is a newly emergent probabilistic logic system that allows for
reasoning with uncertainty. It has been shown to be an effective system for modelling
multi-agent decision making, analyzing bayesian networks, and several other tasks
where uncertainty is rampant.

Despite these benefits, the tooling landscape surrounding Subjective Logic is very
desolate. If one wishes to utilize SL for their own uncertain reasoning applications, it
would be beneficial to be able to utilize an off-the-shelf library of operators, to avoid
rewriting them each and every time. Unfortunately to the best of the author's knowledge,
such a library does not yet exist.

To counter this, we have begun development of a \emph{compositional Subjective Logic
operator library} that can be used to validate and evaluate well-formed expressions in SL.
We utilize \emph{monads}, an abstraction popular in the \emph{functional programming}
community in order to retain compositionality of our operators while also offering
memoization and run-time error message propogation to the users of our library.

\section{Programming with Monads}



\section{A Subjective Logic Implementation}


\section{An Example Application}


%\section{Future Work}


\section{Conclusions}



\end{document}
